Using Scagnostics to find Interesting Projections of Multivariate Data

### 2018-09-11 scagnostics.Rmd
More optimisation, testing change in index when moving into V5 vs V6.

We next compare the various indexvalues for three views: V1 vs V2, V5 vs V6 and V1 vs V6 (capturing differences from the different distribution in V6).

@@

```

@@

@@

### Looking down the pipe
TIC should be able to uncover special view, compare performance of all indices on guided tour path. (also cmass and holes?)
Not currently finding "best" view. Options: compare results from n different random starting planes, try with smaller space (e.g. omitting variable 1 and 2), turn some of the guided tour parameters
```{r findpipe, fig.width=6, fig.height=7, fig.cap="Guided tour optimising TIC index for Pipe 1000 dataset."}
set.seed(2018)
pipeResc <- rescale(pipe1000)
pipeTour <- save_history(pipeResc,
                         guided_tour(mineIndex("TIC"), n=10, cooling = 0.8), start = p58)
pipeTourFull <- as.list(interpolate(pipeTour))
pipeTourRes <- getProj(pipe1000, pipeTourFull, "Pipe", 1000)
pipeTourResMelt <- melt(pipeTourRes, id=c("t", "name", "size"))
ggplot(pipeTourResMelt, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

```{r testpipe}
iLast <- length(pipeTourFull)
fProj <- pipeTourFull[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

### Finding sine waves
Use spline2d to try and find sine shape, compare performance of all indices on guided tour path.

>>>>

We next compare the various indexvalues for three views: V1 vs V2, V5 vs V6 and V1 vs V6 (capturing differences from the different distribution in V6). For the Sphere distribution by construction we see only random fluctuations between the different views, which can be sizable in particular for small datasets. Considering next the Pipe distribution, we see that skinny, stringy and dcor2d are increased for the special view, but in particular MIC and TIC index are much larger and may be used to detect this type of structure in the dataset. Finally the Sine distribution shows that several of the indices are maximised in the special view, in particular also the splines2d index designed for the detection of fuctional dependence.

@@

@@

## Transition
We repeat the same exercise, but now we are interested in how the index values change when going from the view in V1 vs V2 to the view in V5 vs V6.
```{r plannedtourpath}
m1 <- matrix(c(1,0,0,0,0,0,0,1,0,0,0,0),ncol=2)
m2 <- matrix(c(0,0,0,0,1,0,0,0,0,0,0,1),ncol=2)
#this is silly but seems that first two entries are being ignored, so need some fake entries
m3 <- matrix(c(1,0,0,0,1,0,0,1,0,0,0,1),ncol=2)
m4 <- matrix(c(1,1,1,0,0,0,0,1,1,0,0,0),ncol=2)
t2 <- save_history(sin100,tour_path=planned_tour(list(m3,m4,m1,m2)))
t2full <- as.list(interpolate(t2))
fullResPlanned <- getProj(sin100, t2full, "sin", 100) %>%
  rbind(getProj(sin1000, t2full, "sin", 1000)) %>%
  rbind(getProj(sphere100, t2full, "sphere", 100)) %>%
  rbind(getProj(sphere1000, t2full, "sphere", 1000)) %>%
  rbind(getProj(pipe100, t2full, "pipe", 100)) %>%
  rbind(getProj(pipe1000, t2full, "pipe", 1000))
```

```{r plannedgrandtour, fig.width=6, fig.height=7, fig.cap="Short tour paths with scagnostics computed on projections. Most of the indices exhibit sharp jumps in scagnostic value."}
fullResMeltPlanned <- melt(fullResPlanned, id=c("t", "name", "size"))
ggplot(fullResMeltPlanned, aes(x=t, y=value)) +
  geom_line(aes(color=name, linetype=factor(size))) + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

We can see a pretty clear and smooth increase in dcor2d and splines2d as well as MIC and TIC index when approaching the special view of the Sine distribution, which should therefore be efficient to optimise. On the other hand while we also see smooth increase in MIC and TIC when approaching the special view of the Pipe distribution, the increase only becomes apparant much closer to the final projection, making this structure more challenging to detect with a guided tour.
(What does this mean for optimisation? How can I detect such "smaller scale increase"? Maybe try "search_better" instead of search_geodesic? increase alpha and decrease cooling? increase max.tries?)
Can force it when using search_better with large alpha, cooling=1, max.tries=1000, but not ideal..
Difficulties also from "fake" structure from the sampling boundaries?
Possible issue with search_geodesic is that the search window is fixed in the code..

@@

### Looking down the pipe
As seen before (reference figure) MIC or TIC should be able to find the view down the pipe. The optimisation itself is however expected to be challenging because one has to be fairly close to the optimal projection to observe the increase in index value. We therefore suggest a two step procedure for the uncovering of such "small scale structures". The first step consists in "scouting" projections. Here we use the "search_better" optimisation available with the tourr package, with a standard search window $\alpha = 0.5$ and disabeling the cooling (in practice this can be done by fixing the cooling parameter to $1$), together with a large number of tries (setting max.tries to 5000 here). The resulting search will quickly steer away from the most unintresting projections, followed by a broad scan of the thus identified region. While this does not maximise the index function, it provides an informed starting plane that can be used in a second step where we use the "search_geodesic" optimisatino to improve the optimisation.
The described procedure is CPU time intensive, however it is promising that we manage to uncover such small scale structure in the distribution. On the other hand no "guarantee" can be provided for uncovering the special view. Parameters for the scouting phase will need to be adapted to different examples, and with increasing number of dimensions it may be better to consider several runs with different starting projections.
This gives some indication on how to handle difficult datasets. As we will see below most of the time standard "search_geodesic" optimisation will be efficient enough to uncover the interesting views.

```{r findpipe, fig.width=6, fig.height=7, fig.cap="Guided tour optimising TIC index for Pipe 1000 dataset.", echo=FALSE}
set.seed(2018)
pipeResc <- rescale(pipe1000) #unselect 2 variables to make optimisation easier
pipeTour <- save_history(pipeResc,
                         guided_tour(mineIndex("TIC"), search_f = tourr:::search_better,
                                     cooling=1, alpha = 0.5, max.tries = 5000))
pipeTourFull <- as.list(interpolate(pipeTour))
pipeTourRes <- getProj(pipe1000, pipeTourFull, "Pipe", 1000)
pipeTourResMelt <- melt(pipeTourRes, id=c("t", "name", "size"))
ggplot(pipeTourResMelt, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

```{r testpipe}

iLast <- length(pipeTourFull)
fProj <- pipeTourFull[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

```{r refine}
pipeTour2 <- save_history(pipeResc, guided_tour(mineIndex("TIC")), start = fProj)
pipeTourFull2 <- as.list(interpolate(pipeTour2))
pipeTourRes2 <- getProj(pipe1000, pipeTourFull2, "Pipe", 1000)
pipeTourResMelt2 <- melt(pipeTourRes2, id=c("t", "name", "size"))
ggplot(pipeTourResMelt2, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

```{r testpipe2}
iLast <- length(pipeTourFull)
fProj <- pipeTourFull[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

### Finding sine waves
```{r findsine, fig.width=6, fig.height=7, fig.cap="Guided tour optimising TIC index for Sine 1000 dataset.", echo=FALSE}
set.seed(2018)
sineResc <- rescale(sin1000) #unselect 2 variables to make optimisation easier
sineTour <- save_history(sineResc,
                         guided_tour(splineIndex()))
sineTourFull <- as.list(interpolate(sineTour))
sineTourRes <- getProj(sin1000, sineTourFull, "Sine", 1000)
sineTourResMelt <- melt(sineTourRes, id=c("t", "name", "size"))
ggplot(sineTourResMelt, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

```{r testsine}
iLast <- length(sineTourFull)
fProj <- sineTourFull[[iLast]]
dProj <- as.tibble(sineResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```
